sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-01-24 doc/appendix-pcod-VAST-poisson-link.Rmd
Updating paper

```{r settings, echo = TRUE}
# get coordinates in geographic coordinates from UTM projection, after changing units to m
qcs_grid_ll <- qcs_grid
qcs_grid_ll$Y <- qcs_grid_ll$Y * 1000
qcs_grid_ll$X <- qcs_grid_ll$X * 1000

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
ind <- get_index(p, bias_correct = FALSE, area = rep(4, nrow(qcs_grid)))
tictoc::toc()
```

>>>>

```{r settings, echo = TRUE}
# get coordinates in geographic coordinates from UTM projection, after changing units to m
qcs_grid_ll <- replicate(qcs_grid, "year", unique(pcod$year))
qcs_grid_ll$Y <- qcs_grid_ll$Y * 1000
qcs_grid_ll$X <- qcs_grid_ll$X * 1000

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
survey_grid <- replicate_df(qcs_grid, "year", unique(pcod$year))
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = survey_grid, return_tmb_object = TRUE)
ind <- get_index(p, bias_correct = FALSE, area = 4)
tictoc::toc()
```

### 2023-01-24 doc/appendix-pcod-VAST-tweedie.Rmd
Print sdmTMB model
Updating paper

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
ind_sdmTMB <- get_index(pred, area = 4)
```

>>>>

```{r sdmTMB-example-print, echo = TRUE}
fit_sdmTMB
```

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
survey_grid <- replicate_df(qcs_grid, "year", unique(pcod$year))
pred <- predict(fit_sdmTMB, newdata = survey_grid, return_tmb_object = TRUE)
ind_sdmTMB <- get_index(pred, area = 4)
```

### 2023-01-24 doc/paper-jss/inla-comparison.Rmd
Edits
Updating paper

@@

```{r sdm-predictions}
nd <- filter(qcs_grid, year >= 2007)
pred <- predict(m, newdata = nd)
```

>>>>

```{r inla-knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  dev = "pdf"
)
```

@@

```{r sdm-predictions}
nd <- replicate_df(qcs_grid, time_name = "year", time_values = unique(d$year))
pred <- predict(m, newdata = nd)
```

### 2023-01-24 doc/paper-jss/owls.Rmd
Updating paper

```{r rds1, echo=FALSE, cache=FALSE}
rds_saved <- file.exists(here::here("data/snow_w_main_effect_0_150km.rds"))
rds_saved2 <- file.exists(here::here("data/snow_w_main_effect_0_150km2.rds"))
```

```{r fit-negb2, echo=TRUE, eval=!rds_saved, warning=FALSE}
fit_owl <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh,
  reml = TRUE,
  silent = FALSE
)
```

```{r cache-negb2, echo=FALSE, eval=!rds_saved}
saveRDS(fit_owl, here::here("data/snow_w_main_effect_0_150km.rds"))
```

```{r read-negb2, echo=FALSE, eval=rds_saved}
fit_owl <- readRDS(here::here("data/snow_w_main_effect_0_150km.rds"))
```

@@

```{r owl-newton, results='hide', eval=!rds_saved2}
fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
```

```{r cache-negb2-2, echo=FALSE, eval=!rds_saved2}
saveRDS(fit_owl2, here::here("data/snow_w_main_effect_0_150km2.rds"))
```

```{r read-negb2-2, echo=FALSE, eval=rds_saved2}
fit_owl2 <- readRDS(here::here("data/snow_w_main_effect_0_150km2.rds"))
```

>>>>

```{r fit-negb2, echo=TRUE, warning=FALSE}
fit_owl <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh,
  reml = TRUE,
  silent = FALSE
)
```

@@

```{r owl-newton, results='hide'}
fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
```

### 2023-01-24 doc/paper-jss/paper-jss.Rmd
Updating paper

Speed-wise, \pkg{sdmTMB} (and due to the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
<!-- Furthermore, although both \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \proglang{R}-\pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores. -->

@@

Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
The included \pkg{TMB} `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
<!-- Not sure what kind of refs you want here?  \proglang{R}-\pkg{INLA}'s non-gaussian random fields are mentioned here @bakka2018review-->
<!-- \pkg{sdmTMB} does not currently include  -->
<!-- We may introduce explicit delta models, but many such models can already be fit with \pkg{sdmTMB} by modelling the two components separately. -->
<!-- We may introduce capability for multivariate response data, but \pkg{VAST} already focuses on such models with a \pkg{TMB} backend. -->
<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
<!-- Line 275: I would suggest adding a MAJOR benefit of \pkg{sdmTMB} for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates! This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it. It saves me a bunch of time. If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it. -->

<!-- Spatially and spatiotemporally explicit data are increasingly collected and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019]. -->
<!-- These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate. -->
<!-- We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users. -->

@@

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \proglang{R}-\pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.

>>>>

With respect to computational speed, \pkg{sdmTMB} (and due to having the same \pkg{TMB} backend, \pkg{VAST}) were fastest up to at least 1000 mesh nodes at approximately 4--16 times faster than \proglang{R}-\pkg{INLA}/\pkg{inlabru} across the range of mesh complexities considered here.
These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
However, for users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \proglang{R}-\pkg{INLA}/\pkg{inlabru} is likely to be considerably faster than passing the same model from \pkg{sdmTMB}/\pkg{VAST} to \pkg{tmbstan} for full Bayesian inference.
<!-- Furthermore, although both \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{sdmTMB} can use parallel processing, the PARDISO library [@pardiso-7.2c] within \proglang{R}-\pkg{INLA}/\pkg{inlabru} allows memory to be shared across cores. -->

@@

Additional functionality in \pkg{sdmTMB} not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
Future development may include additional zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous-time models [e.g., @blangiardo2015book], multivariate responses, and non-Gaussian random fields [e.g., @anderson2019].
The included \pkg{TMB} `.cpp` file provides a high-quality and tested model template that can be modified to add additional features.

@@

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \proglang{R}-\pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, M. Martin, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.

### 2023-01-24 doc/paper-jss/pcod.Rmd
Updating paper

Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year. There is a grid built into the \pkg{sdmTMB} package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above. In this case, that is a `year` column as well as `X` and `Y`.

```{r predict-newdata}
pred_qcs <- predict(fit, newdata = qcs_grid)
```

@@

```{r pcod-sims, cache=TRUE}
pred_sims <- predict(fit, newdata = qcs_grid, nsim = 200)
dim(pred_sims)
```

@@

```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='ht', fig.align='center'}
qcs_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
ggplot(qcs_grid, aes(X, Y, fill = cv)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10", option = "D")
```

@@

```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='ht', fig.align='center'}
qcs_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = qcs_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = qcs_grid$area, bias_correct = TRUE)

>>>>

Next, we want to predict on a fine-scale grid over the entire survey domain so we can sum up the expected biomass across the entire region each year. There is a grid built into the \pkg{sdmTMB} package for Queen Charlotte Sound (named `qcs_grid`). Our prediction grid also needs to have all the covariates that we used in the model above. In this case, that is a `year` column as well as `X` and `Y`. We will replicate the internal spatial grid over all years (calling it `survey_grid`) and predict on it.

```{r predict-newdata}
survey_grid <- replicate_df(
  qcs_grid, 
  time_name = "year", 
  time_values = unique(pcod$year)
)

pred_qcs <- predict(fit, newdata = survey_grid)
```

@@

```{r pcod-sims, cache=TRUE}
pred_sims <- predict(fit, newdata = survey_grid, nsim = 200)
dim(pred_sims)
```

@@

```{r sims-cv-plot, cache=TRUE, fig.cap="Spatiotemporal coefficient of variation.", fig.pos='ht', fig.align='center'}
survey_grid$cv <- apply(pred_sims, 1, function(x) sd(exp(x)) / mean(exp(x)))
ggplot(survey_grid, aes(X, Y, fill = cv)) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10", option = "D")
```

@@

```{r pcod-app-index, fig.asp=0.45, out.width="5in", cache=TRUE, fig.cap="Geostatistical population index.", fig.pos='ht', fig.align='center'}
survey_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = survey_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = survey_grid$area, bias_correct = TRUE)

### 2023-01-24 doc/paper-jss/timing.Rmd
Edits
Updating paper

\clearpage
# Speed testing related packages {#app:speed}

@@

```{r sim-timing, echo=TRUE}
set.seed(1)
n_obs <- 1000
predictor_dat <- data.frame(
  X = runif(n_obs), Y = runif(n_obs),
  a1 = rnorm(n_obs)
)
mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)

@@

```{r mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE, cache=TRUE}
# -------------------------------------------------------------------------
# Code in this chunk is from:
#
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
smooth.construct.spde.smooth.spec <- function(object, data, knots) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  if (is.null(object$xt)) {
    if (dim == 1) {
      t <- seq(min(x), max(x), len = object$bs.dim)
      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
    } else {
      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
    }
  } else {
    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
    mesh <- object$xt$mesh
  }
  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
  inlamats <- INLA::inla.mesh.fem(mesh)
  object$S <- list()
  object$S[[1]] <- as.matrix(inlamats$c1)
  object$S[[2]] <- 2 * as.matrix(inlamats$g1)
  object$S[[3]] <- as.matrix(inlamats$g2)
  object$L <- matrix(c(2, 2, 2, 4, 2, 0), ncol = 2)
  object$rank <- rep(object$bs.dim, 3)
  object$null.space.dim <- 0
  object$mesh <- mesh
  object$df <- ncol(object$X)
  class(object) <- "spde.smooth"
  return(object)
}
Predict.matrix.spde.smooth <- function(object, data) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  Xp <- INLA::inla.spde.make.A(object$mesh, x)
  return(as.matrix(Xp))
}
# End of code from
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
# -------------------------------------------------------------------------
```

```{r mgcvfit-timing, echo=TRUE, eval=FALSE}
# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth()
# from supplement of:
# Miller, D.L., Glennie, R. & Seaton, A.E. (2019). Understanding the Stochastic
# Partial Differential Equation approach to smoothing. Journal of Agricultural,
# Biological and Environmental Statistics.
# https://doi.org/10.1007/s13253-019-00377-z

>>>>

# Speed testing related packages {#app:speed}

@@

```{r sim-timing, echo=TRUE}
set.seed(123)
n_obs <- 1000
predictor_dat <- data.frame(
  X = runif(n_obs), Y = runif(n_obs),
  a1 = rnorm(n_obs)
)
mesh_sdmTMB <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), mesh = mesh)

@@

```{r mgcv-spde-funcs-timing, echo=FALSE, eval=TRUE}
# -------------------------------------------------------------------------
# Code in this chunk is from:
#
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
smooth.construct.spde.smooth.spec <- function(object, data, knots) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  if (is.null(object$xt)) {
    if (dim == 1) {
      t <- seq(min(x), max(x), len = object$bs.dim)
      mesh <- INLA::inla.mesh.1d(loc = t, degree = 2, boundary = "free")
    } else {
      stop("For 2D, mesh must be supplied as argument xt$mesh in s(...,xt = )")
    }
  } else {
    if (class(object$xt$mesh) != "inla.mesh") stop("xt must be NULL or an inla.mesh object")
    mesh <- object$xt$mesh
  }
  object$X <- as.matrix(INLA::inla.spde.make.A(mesh, x))
  inlamats <- INLA::inla.mesh.fem(mesh)
  object$S <- list()
  object$S[[1]] <- as.matrix(inlamats$c1)
  object$S[[2]] <- 2 * as.matrix(inlamats$g1)
  object$S[[3]] <- as.matrix(inlamats$g2)
  object$L <- matrix(c(2, 2, 2, 4, 2, 0), ncol = 2)
  object$rank <- rep(object$bs.dim, 3)
  object$null.space.dim <- 0
  object$mesh <- mesh
  object$df <- ncol(object$X)
  class(object) <- "spde.smooth"
  return(object)
}
Predict.matrix.spde.smooth <- function(object, data) {
  dim <- length(object$term)
  if (dim > 2 | dim < 1) stop("SPDE Matern can only be fit in 1D or 2D.")
  if (dim == 1) {
    x <- data[[object$term]]
  } else {
    x <- matrix(0, nr = length(data[[1]]), nc = 2)
    x[, 1] <- data[[object$term[1]]]
    x[, 2] <- data[[object$term[2]]]
  }
  Xp <- INLA::inla.spde.make.A(object$mesh, x)
  return(as.matrix(Xp))
}
# End of code from
# Miller, D.L., Glennie, R. & Seaton, A.E. Understanding the Stochastic Partial
# Differential Equation Approach to Smoothing. JABES 25, 1–16 (2020).
# https://doi.org/10.1007/s13253-019-00377-z
#
# Re-used here under a Creative Commons Attribution 4.0 International License
# http://creativecommons.org/licenses/by/4.0/
# -------------------------------------------------------------------------
```

```{r mgcvfit-timing, echo=TRUE, eval=TRUE}
# define smooth.construct.spde.smooth.spec() and Predict.matrix.spde.smooth()
# from supplement of:
# Miller, D.L., Glennie, R. & Seaton, A.E. (2019). Understanding the Stochastic
# Partial Differential Equation approach to smoothing. Journal of Agricultural,
# Biological and Environmental Statistics.
# https://doi.org/10.1007/s13253-019-00377-z
